RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2015-01-01 manuscripts/manuscript.Rmd
Rewrite of rest of custom metadata section.

This URL links us to a web page providing both a human and machine-readable
version of precisely what the term `title` means, showing us that it
refers to `title` in the sense of a creative work, rather than the title
of a person, a legal title, the verb title, etc.  This precision can allow
us to perform precise computer reasoning on metadata, without relying on
heuristics or assumptions, as we illustrate later.  For the moment though,
the need to refer to an explicit controlled vocabulary (or ontology)
with a url, saying "dc:title" rather than just saying the property is
"title", is something of a burden in creating metadata.  The previous
examples involve functions that already know what kind of metadata is
being provided (a title, a data creator, etc), and so can provide the
appropriate vocabulary automatically.  The main advantage of linking to
an external vocabulary, however, is the ability to express concepts that
are not already "built in" to the data model.  `RNeXML` provides tools
to generate arbitrary metadata annotations given an external vocabulary,
such as we illustrate here.

From the list of namespaces in `prefixes`, we can follow links to a handful
of established vocabularies and find terms that are not included in
`add_basic_meta()`.  For instance, by resolving the location identified
by the `prism` prefix,

```{r}
prefixes["prism"]
```

we will find that there is a term for `modificationDate` for a file.  We create
a `meta` element containing this annotation using the `meta` function:

@@

We can add this annotation to our existing `birds` NeXML file using the
`add_meta()` function.  Because we do not specify a level, it is added to
the root node, referring to the NeXML file as a whole.

@@

The built-in vocabularies are just the tip of the iceberg of established
vocabularies. Here we add an annotation from the `skos` namespace which
describes the history of where the data comes from:

```{r}
history <- meta(property = "skos:historyNote",
  content = "Mapped from the bird.orders data in the ape package using RNeXML")
```

Because `skos` is not in the current namespace list, we add it with a
url when adding this meta element.  We also specify that this annotation
be placed at the level of the `trees` sub-node in the NeXML file.

```{r}
birds <- add_meta(history, 
                birds, 
                level = "trees",
                namespaces = c(skos = "http://www.w3.org/2004/02/skos/core#"))
```

For finer control of the level at which a `meta` element is added,
we will manipulate the `nexml` R object directly using S4 sub-setting,
as shown in the supplement.

Much richer metadata annotation is possible. Later we illustrate how
metadata annotation can be used to extend the base NeXML format to
represent new forms of data while maintaining compatibility with any
NeXML parser. The `RNeXML` package can be easily extended to support
helper functions such as `taxize_nexml` to add additional metadata
without imposing a large burden on the user.

## Reading NeXML metadata

A call to the `nexml` object prints some metadata summarizing the data structure:

```{r }
birds
```

We can extract all metadata pertaining to the NeXML document as a whole
(annotations of the XML root node, `<nexml>`) with the command

This returns a named list of available metadata. We can see the kinds
of metadata recorded from the names (showing the first 4):

```{r}
names(meta)[1:4]
```

and can ask for a particular element using the standard list sub-setting
mechanism (i.e. either the name of an element or its numeric position),

```{r}
meta[["dc:title"]]
```

All metadata terms must belong to an explicit *namespace* or vocabulary
that allows a computer to interpret the term precisely. The prefix (before
the `:`) indicates to which vocabulary the term belongs, e.g. `dc` in
this case.  The `get_namespaces` function tells us the definition of
the vocabulary using a link:

<!-- the citations run off the page -->
Common metadata can be accessed with a few dedicated functions:

```{r get_taxa}
get_taxa(birds)
```

Which returns text from the otu element labels, typically used to define
taxonomic names, rather than text from explicit meta elements.

We can also access metadata at a specific level (or use `level=all`
to extract all meta elements in a list).  Here we show only the first
few results:

```{r} 
otu_meta <- get_metadata(birds, level="otu")
otu_meta[1:4]
```

This approach does not preserve the relationship between the metadata element and the 
`otu` it was annotating, and as such is most useful for quick visual exploration. 
Metadata values for specific elements or sets of elements can always be accessed using
the S4 sub-setting mechanisms (see section below).

>>>>

The `get_metadata()` method returns, as named list, the metadata
annotations for a given `nexml` object at a given level, with the
whole NeXML document being the default:

The `level` parameter can be used to extract the metadata objects for
the data at specific levels (such as `"otu"` in the following
example), or by specifying `"all"` to extract all metadata objects:

```{r} 
otu_meta <- get_metadata(birds, level="otu")
```

However, because the returned list does not include the data elements
to which the metadata are attached, using the method in this way is
only useful for identifying which metadata are present and their
values. The recommended approach for accessing the metadata attached
to specific data elements is described below. <!-- or is that in the vignette now? -->

A few select metadata properties for which support is built-in have
dedicated accessor methods, for example the citation:

The `meta()` method creates a new metadata object from a poperty name
and content (value). For example, the following creates a modification
date metadata object, using the modificationDate property in the PRISM
vocabulary:

@@

The `add_meta()` function adds metadata objects as annotations to a
`nexml` object at a specified level. The default level is the NeXML document as a whole:

@@

If the prefix used by the metadata property is not among the built-in
ones (which can be obtained using the `get_namespaces()` as described
above), it has to be provided along with its URI as the `namespaces`
parameter. For example, the following uses the SKOS vocabulary (which
currently is not built-in) to add a note to the trees in the `nexml`
object:

```{r}
history <- meta(property = "skos:historyNote",
  content = "Mapped from the bird.orders data in the ape package using RNeXML")
birds <- add_meta(history, 
                birds, 
                level = "trees",
                namespaces = c(skos = "http://www.w3.org/2004/02/skos/core#"))
```

`RNeXML`, by virtue of subsetting the S4 `nexml` object, also offers
fine control of the level at which a `meta` element is added, for
which the package vignette contains examples.
